Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-02 manuscripts/reply/reply-abstract.tex
More editing of reply ms.

Biogeographers frequently seek to explain population and species
differentiation on geographical phenomena.
Establishing that a set of population splitting events occurred
at the same time can be a potentially persuasive argument that a set of taxa
were affected by the same geographic events.
Unfortunately, estimating divergence times precisely when one lacks
precise information about the rate of molecular evolution is a notoriously
difficult problem in evolutionary biology.
\citet{Huang2011} introduced an approximate Bayesian approach (implemented
in the software \msb) which can be used to estimate the probabilities of
models in which multiple sets of taxa diverge at the same time. 
% This method  does not require a common set of character data that links all
% of the taxa under investigation, but does require the relative mutation
% rates among the taxa and loci to be assumed \emph{a priori}.
Recently, \citet{Oaks2012} used this model-choice framework to 
study 22 pairs of vertebrate lineages which are distributed across
the Philippines; they also studied the behavior of the \msb approach
using computer simulations.
\citet{Oaks2012} found that the model was very sensitive to the prior and
that the inference method had low power to detect variation in divergences
times.
These results were not surprising, in light of a rich statistical literature
showing that the marginal likelihood of a model is sensitive to the
use of vague priors.
Because this sensitivity to prior assumptions affects the crucial insights 
that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
that users of the approach should carefully assess the robustness of their 
conclusions to different priors.
According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
was due to excessively broad priors on divergence times, leading to 
inadequate numbers of simulation replicates.
They proposed a new model-averaging approach that uses narrow, empirically
informed uniform priors.
Here we demonstrate that the approach of \citet{Hickerson2013} is
dangerous in the sense that the empirically-derived priors often
exclude from consideration the true values of the models' parameters.
On a more fundamental level, we question the value of adopting an empirical
Bayesian stance for this model-choice problem, because it can mislead model
posterior probabilities, which are inherently measures of belief in the
models after prior knowledge is updated by the data.
The robust Bayesian approach of conducting analyses under a variety
of priors can reveal prior sensitivity and communicate which assumptions
underlie model inference.
Furthermore, simulations provide insight into the temporal resolution of
the method, which in turn helps guide interpretation of results.

>>>>

Biogeographers frequently seek to explain population and species
differentiation on geographical phenomena.
Establishing that a set of population splitting events occurred
at the same time can be a potentially persuasive argument that a set of taxa
were affected by the same geographic events.
Unfortunately, estimating divergence times precisely when one lacks
precise information about the rate of molecular evolution is a notoriously
difficult problem in evolutionary biology.
\citet{Huang2011} introduced an approximate Bayesian approach (implemented
in the software \msb) which can be used to estimate the probabilities of
models in which multiple sets of taxa diverge at the same time. 
% This method  does not require a common set of character data that links all
% of the taxa under investigation, but does require the relative mutation
% rates among the taxa and loci to be assumed \emph{a priori}.
Recently, \citet{Oaks2012} used this model-choice framework to 
study 22 pairs of vertebrate lineages which are distributed across
the Philippines; they also studied the behavior of the \msb approach
using computer simulations.
\citet{Oaks2012} found the method was very sensitive to prior assumptions and
had low power to detect variation in divergence times.
These results were not surprising, in light of a rich statistical literature
showing that the marginal likelihood of a model is sensitive to the
use of vague priors.
Because this sensitivity to prior assumptions affects the crucial insights 
that a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended
that users of the approach should carefully assess the robustness of their 
conclusions to different priors.
According to \citet{Hickerson2013}, the lack of robustness in \msb analyses
was due to excessively broad priors on divergence times, leading to 
inadequate numbers of simulation replicates.
They proposed a new model-averaging approach that uses narrow, empirically
informed uniform priors.
Here we demonstrate that the approach of \citet{Hickerson2013} is
dangerous in the sense that the empirically-derived priors often
exclude from consideration the true values of the models' parameters.
On a more fundamental level, we question the value of adopting an empirical
Bayesian stance for this model-choice problem, because it can mislead model
posterior probabilities, which are inherently measures of belief in the
models after prior knowledge is updated by the data.
The robust Bayesian approach of conducting analyses under a variety
of priors can reveal prior sensitivity and communicate which assumptions
underlie model inference.
Furthermore, simulations provide insight into the temporal resolution of the
method, which helps guide interpretation of results.

### 2014-07-02 manuscripts/reply/reply-main.tex
Working on reply ms.
Fixing two citations of `Oaks2014dpp`; changing `citet` to `citep`.
More editing of reply ms.

\msb \citep{Huang2011} uses approximate Bayesian computation (ABC) to estimate
the distribution of divergence times among co-distributed pairs of taxa.
It approximates a posterior probability over models that range from a single
divergence-time parameter (i.e., simultaneous divergence of all pairs of taxa)
to the fully generalized model in which each pair of taxa diverged at a
unique time.
A full description of the model is given in \citet{Oaks2012}.
Frequently, the exact values of the divergence times are hard 
to determine because researchers typically lack precise knowledge 
of the rate at which substitutions occur.
The exact divergence time is therefore often viewed as a nuisance parameter,
and researchers focus on the number of divergence events.
The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
on the fundamental methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       support for models with few divergence events shared across taxa. Thus,
       the primary inference enabled by the approach is very sensitive to the
       priors on divergence times.
\end{itemize}

This is not surprising given the rich statistical literature that shows that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection
\citep[e.g.,][]{Jeffreys1939,Lindley1957}.
Accordingly, \citet{Oaks2012} suggest the primary cause of spurious support
for models with few divergence parameters is the greater marginal likelihoods
of these models under vague uniform priors, relative to more parameter-rich
models.
Models with more divergence-time parameters integrate over \emph{much} greater
parameter space with low probability of producing the data, yet relatively high
prior density imposed by the uniformly distributed prior on divergence times.
Note, this is a statistical issue, extrinsic to \msb, and does not question the
soundness of the numerical approximation machinery of \msb.

\citet{Hickerson2013} suggest the bias is the result of a numerical issue
intrinsic to \msb, concluding the method's rejection algorithm is inefficient
and will be biased toward models with fewer divergence-time parameters, because
the parameter space of these models are more densely sampled relative to models
with more divergence-time parameters.
We agree that sampling error is present in all numerical Bayesian estimation
methods, however, we show here that it is not a major contributer to the
biases found by \citet{Oaks2012}.
We present results here of analyses that, when combined with the results
of \citet{Oaks2012}, strongly suggest that \msb prefers models with few
divergence-time parameters, because they have greater likelihoods when
averaged over broad uniform priors on divergences times.

@@

In this paper, we discuss the potential theoretical and practical
considerations of using empirically informed priors for Bayesian model choice.
Furthermore, we evaluate the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
\citet{Oaks2012}.
In their re-analysis of the dataset of \citet{Oaks2012}, \citet{Hickerson2013}
made an error by mixing different units of time, which makes the results
presented in their response difficult to interpret (see Supporting Information
for details).
Here, we avoid this error, but still find their approach to be biased toward
clustered divergence models.
Furthermore, the approach provides another means by which the method can
``escape'' models with large parameter space, which manifests in the preference
of models that exclude from consideration the true values of the model's
parameters.
Our results, when combined with those of \citet{Oaks2012}, suggest that it is
difficult to choose a uniformly distributed prior on divergence times that is
broad enough to confidently contain the true values of parameters while being
narrow enough to avoid spurious support of models with less parameter space.
However, recent work shows more flexible probability distributions without a
hard upper bound (e.g., gamma) can accommodate prior uncertainty in divergence
times without inhibiting the marginal likelihoods of models with more
divergence-time parameters, and as a result, increase the method's power to
detect temporal variation in divergences
\citet{Oaks2014dpp}.

@@

The strong preference for the questionable priors on divergence times (their
$M_1$ and $M_2$) in the model-averaged results of \citet{Hickerson2013} suggest
their approach is biased toward models with less parameter space, and as a
consequence, is biased toward estimating model-averaged posteriors dominated by
models that exclude true values of parameters of the model.
We explored this possibility in two ways.
First, we re-analyzed the Philippines dataset using the model-averaging
approach of \citet{Hickerson2013}, but set one of the prior models with a
uniform prior on divergence times that is unrealistically narrow, and almost
certainly excludes most, if not all, of the true divergence times of the 22
taxon pairs.
If small likelihoods of large models cause the method to prefer models with
less parameter space, we expect \msb will preferentially sample from this
incorrect model yielding a posterior that is incorrect (i.e., the
model-averaged posterior will be dominated by an incorrect model that excludes
the truth).
Second, we generated simulated datasets for which the divergence times are
drawn from an exponential distribution and applied the approach of
\citet{Hickerson2013} to each of them to see how often the method excludes the
truth.

@@

Whether or not one chooses to use empirically informed priors, our results
suggest that it is difficult to choose an uniformly distributed prior on
divergence times that is broad enough to confidently contain the true values of
parameters while being narrow enough to avoid bias toward models with less
parameter space.
While we do not assume ``that all previous \msb results are invalid,'' as
suggested by \citet{Hickerson2013}, we do conservatively recommend that the
common inference of temporally clustered divergences \citep{Barber2010,
    Bell2012, Carnaval2009, Chan2011, Daza2010, Hickerson2006, Huang2011,
Lawson2010, Leache2007, Plouviez2009, Stone2012, Voje2009}, when not
accompanied with the necessary analyses to assess the robustness and temporal
resolution of such results, should be treated with caution, because the method
has been shown to spuriously infer clustered divergences over a range of prior
conditions.
Fortunately, alternative prior probability distributions allow prior
uncertainty to be accommodated while avoiding excessive prior density in
regions of low likelihood, which greatly improves the power of the method
\citet{Oaks2014dpp}.

>>>>

\msb \citep{Huang2011} uses approximate Bayesian computation (ABC) to estimate
the distribution of divergence times among co-distributed pairs of taxa.
It approximates a posterior probability over models that range from a single
divergence-time parameter (i.e., simultaneous divergence of all pairs of taxa)
to the fully generalized model in which each pair of taxa diverged at a
unique time.
% A full description of the model is given in \citet{Oaks2012}.
% Frequently, the exact values of the divergence times are hard 
% to determine because researchers typically lack precise knowledge 
% of the rate at which substitutions occur.
% The exact divergence time is therefore often viewed as a nuisance parameter,
% and researchers focus on the number of divergence events.
The papers of \citet{Oaks2012} and \citet{Hickerson2013} are in agreement
on the fundamental methodological point about the model selection performed in \msb:
\begin{itemize}
   \item The use of vague priors on the divergence-time parameters can lead to
       support for models with few divergence events shared across taxa. Thus,
       the primary inference enabled by the approach is very sensitive to the
       priors on divergence times.
\end{itemize}

This is not surprising given the rich statistical literature that shows that
marginal likelihoods are very sensitive to the priors used in Bayesian model
selection
\citep[e.g.,][]{Jeffreys1939,Lindley1957}.
Accordingly, \citet{Oaks2012} suggest the primary cause of spurious support
for models with few divergence parameters is the greater marginal likelihoods
of these models under vague uniform priors, relative to more parameter-rich
models.
Models with more divergence-time parameters integrate over \emph{much} greater
parameter space with low probability of producing the data, yet relatively high
prior density imposed by the uniformly distributed prior on divergence times.
% By using such priors on divergence times, we are essentially telling the
% method to favor models with few divergence times even if they do not
% explain the data very well.
Note, this is a statistical issue, extrinsic to \msb, and does not question the
soundness of the numerical approximation machinery of \msb.

\citet{Hickerson2013} suggest the problem is the result of a numerical issue
intrinsic to \msb, concluding the method's rejection algorithm is inefficient
and will be biased toward models with fewer divergence-time parameters, because
the parameter space of these models are more densely sampled relative to models
with more divergence-time parameters.
We agree that sampling error is present in all numerical Bayesian estimation
methods, however, we show here that it is not a major contributer to the
biases found by \citet{Oaks2012}.
We present results here of analyses that, when combined with the results
of \citet{Oaks2012}, strongly suggest that \msb prefers models with few
divergence-time parameters, because they have greater likelihoods when
averaged over broad uniform priors on divergences times.

@@

In this paper, we discuss the potential theoretical and practical
considerations of using empirically informed priors for Bayesian model choice.
Furthermore, we evaluate the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} as a potential solution to the biases of \msb revealed by
\citet{Oaks2012}.
In their re-analysis of the dataset of \citet{Oaks2012}, \citet{Hickerson2013}
made an error by mixing different units of time, which makes the results
presented in their response difficult to interpret (see Supporting Information
for details).
Here, we avoid this error, but still find their approach to be biased toward
clustered divergence models.
Furthermore, the approach provides another means by which the method can
``escape'' models with large parameter space, which manifests in the preference
of models that exclude from consideration the true values of the model's
parameters.
Our results, when combined with those of \citet{Oaks2012}, suggest that it is
difficult to choose a uniformly distributed prior on divergence times that is
broad enough to confidently contain the true values of parameters while being
narrow enough to avoid spurious support of models with less parameter space.
However, recent work shows more flexible probability distributions without a
hard upper bound (e.g., gamma) can accommodate prior uncertainty in divergence
times without inhibiting the marginal likelihoods of models with more
divergence-time parameters, and as a result, increase the method's power to
detect temporal variation in divergences
\citep{Oaks2014dpp}.

@@

The strong preference for the questionable priors on divergence times (their
$M_1$ and $M_2$) in the model-averaged results of \citet{Hickerson2013} suggest
their approach is biased toward models with less parameter space, and as a
consequence, will estimate model-averaged posteriors dominated by models that
exclude true values of parameters of the model.
We explored this possibility in two ways.
First, we re-analyzed the Philippines dataset using the model-averaging
approach of \citet{Hickerson2013}, but set one of the prior models with a
uniform prior on divergence times that is unrealistically narrow, and almost
certainly excludes most, if not all, of the true divergence times of the 22
taxon pairs.
If small likelihoods of large models cause the method to prefer models with
less parameter space, we expect \msb will preferentially sample from this
incorrect model yielding a posterior that is incorrect (i.e., the
model-averaged posterior will be dominated by an incorrect model that excludes
the truth).
Second, we generated simulated datasets for which the divergence times are
drawn from an exponential distribution and applied the approach of
\citet{Hickerson2013} to each of them to see how often the method excludes the
truth.

@@

Whether or not one chooses to use empirically informed priors, our results
suggest that it is difficult to choose an uniformly distributed prior on
divergence times that is broad enough to confidently contain the true values of
parameters while being narrow enough to avoid bias toward models with less
parameter space.
While we do not assume ``that all previous \msb results are invalid,'' as
suggested by \citet{Hickerson2013}, we do conservatively recommend that the
common inference of temporally clustered divergences \citep{Barber2010,
    Bell2012, Carnaval2009, Chan2011, Daza2010, Hickerson2006, Huang2011,
Lawson2010, Leache2007, Plouviez2009, Stone2012, Voje2009}, when not
accompanied with the necessary analyses to assess the robustness and temporal
resolution of such results, should be treated with caution, because the method
has been shown to spuriously infer clustered divergences over a range of prior
conditions.
Fortunately, alternative prior probability distributions allow prior
uncertainty to be accommodated while avoiding excessive prior density in
regions of low likelihood, which greatly improves the power of the method
\citep{Oaks2014dpp}.
